Living on the edge: crop boundary effects on precision crop yields in Alberta, Canada

### 2021-10-22 Manuscript/manuscript.Rmd
Added appendix with filter

```{r hypotheses, echo=FALSE, fig.cap='Potential yield patterns, depending on ecosystem service effects with distance. Scenario 1: no effect of boundary. Scenario 2: negative effect on yield. Scenario 3: positive effect on yield. Scenario 4: edge effects are shown in red, ecosystem service effect shown blue, leading to an intermediate peak in yield.', out.width='100%'}  
  include_graphics('../Figures/ExamplePlots/hypotheses.png',dpi=NA)
```

@@

Yield data was collected in rectangles of the same length as the data interval (distance = combine ground speed $\times$ interval, typically 1 second) and the same width as the combine header (5-7 m, adjusted by the combine to account for overlap with adjacent swaths).
We used dry yield (tonnes of seed per hectare minus - crop moisture) as our measure of yield, using the size of each rectangle (m$^2$), dry yield (tonnes), and the spatial location.
Because of the large number of yield rectangles per field (30-800 thousand), we used the centroid of each polygon as its location, treating areal data as point data.
Seeding and application rates were constant across fields, so we did not consider inputs in our analysis.
Precision yield data can be highly variable and is prone to extreme outliers, especially when combine ground speeds are low and at the beginning of rows [@griffin2007; @whelan2013], or when the combine is accelerating or decelerating [@arslan2002].
Therefore, we used the following filtering procedure to clean the data prior to model fitting:

@@

Pea crops were less well-represented in our data (N = `r sum(with(datSource,use&modelComplete2&crop=='Peas'))` fields, compared to ), and the effects of boundaries tended to be less consistent. 
Average pea yields (Figure \@ref(fig:peaPlot), top row) were roughly 4.25 T/ha at the field boundary and increased to just below 5 T/ha away from shelterbelt, standard, and other crop boundaries, while the effect of bare, grassland, and wetland boundaries was inconsistent.
Conversely, yield variability (Figure \@ref(fig:peaPlot), bottom row) decreased with distance from grassland, shelterbelt, standard, and wetland boundaries, while the pattern with distance from other crop and bare was inconsistent. 
Unlike canola and wheat crops, there was no obvious evidence of an intermediate increase in yield with distance from the field boundary.

@@

At the level of individual fields, both average and variability of yield changed with distance from the field boundary, and there were strong spatial relationships within each field, likely related to underlying soil or moisture conditions.
Yield tended to increase with distance from field boundaries before plateauing, and there was some evidence of intermediate boosts in yield from shelterbelts next to canola and wheat.
This increase tended to be small in magnitude in canola crops, meaning that ecosystem services likely increased or did not vary with distance from field boundaries (Scenario 2, Figure \@ref(fig:hypotheses)).
There was a fairly large (0.75 T/ha) intermediate increase in average wheat yields away from shelterbelts, indicating a potential reduction in ecosystem services with distance.
However, there was a similar (albeit weaker) pattern associated with bare ground, so this pattern may be caused by something other than the presence of shelterbelts.
This indicates that shelterbelts may have a positive effect on wheat yields, and future work should expand this across a larger geographic domain in order to inform growers about how best to manage the landscapes surrounding their fields.

<!-- For example, average canola yield was about 0.5 T/ha lower close to Standard field edges, and yield SD was about 0.3 T/ha SD higher. -->

Why did we find few consistent effects of boundary type on crop yield, aside from shelterbelts?
Our work examined only the first and last links in the causal chain _Field boundary_ $\rightarrow$ _Insect Abundance_ $\rightarrow$ _Crop Yield_, meaning that are two mediating relationships where the overall relationship could be broken or obscured.
First, field boundaries may not have any influence on insect abundance if separated from the larger landscape context, or spillover of beneficial organisms may be low.
Second, beneficial insect abundance may not influence crop yield if pest pressure or pollination deficits are low, which may vary yearly, or by location.
<!-- However, our work does show that precision yield data could easily be joined to data on insect abundance to model this more completely. -->
Our work suggests that we need to better examine the context in which a) beneficial organisms spillover from field boundaries and b) when organisms in crops provide tangible ecosystem services.
We consider these reasons below:

>>>>

```{r hypotheses, echo=FALSE, fig.cap='Potential yield patterns, depending on ecosystem service effects with distance. Scenario 1: no effect of boundary. Scenario 2: negative edge effect on yield. Scenario 3: positive ecosystem service effect on yield. Scenario 4: edge effects are shown in red, ecosystem service effect shown blue, leading to an intermediate peak in yield.', out.width='100%'}  
  include_graphics('../Figures/ExamplePlots/hypotheses.png',dpi=NA)
```

@@

Yield data was collected in rectangles of the same length as the data interval (distance = combine ground speed $\times$ interval, typically 1 second) and the same width as the combine header (5-7 m, adjusted by the combine to account for overlap with adjacent swaths).
We used dry yield (tonnes of seed per hectare minus - crop moisture) as our measure of yield, using the size of each rectangle (m$^2$), dry yield (tonnes), and the spatial location.
Because of the large number of yield rectangles per field (30-800 thousand), we used the centroid of each polygon as its location, treating areal data as point data.
Seeding and application rates were constant across fields, so we did not consider inputs in our analysis.
Precision yield data can be highly variable and is prone to extreme outliers, especially when combine ground speeds are low and at the beginning of rows [@griffin2007; @whelan2013], or when the combine is accelerating or decelerating [@arslan2002].
Therefore, we used the following filtering procedure to clean the data prior to model fitting (see Appendix A for R code):

@@

Pea crops were less well-represented in our data (N = `r sum(with(datSource,use&modelComplete2&crop=='Peas'))` fields, compared to ), and therefore the effects of boundaries tended to be less consistent. 
Average pea yields (Figure \@ref(fig:peaPlot), top row) were roughly 4.25 T/ha at the field boundary and increased to just below 5 T/ha away from shelterbelt, standard, and other crop boundaries, while the effect of bare, grassland, and wetland boundaries was inconsistent.
Conversely, yield variability (Figure \@ref(fig:peaPlot), bottom row) decreased with distance from grassland, shelterbelt, standard, and wetland boundaries, while the pattern with distance from other crop and bare was inconsistent. 
Unlike canola and wheat crops, there was no obvious evidence of an intermediate increase in yield with distance from the field boundary.

@@

Within individual fields, both the average and variability of yield changed with distance from the field boundary, and there were strong spatial relationships within each field, likely related to underlying soil or moisture conditions.
Mean yield tended to increase with distance from field boundaries before plateauing, and there was some evidence of intermediate boosts in yield from shelterbelts next to canola and wheat, while peas showed no evidence of this.
However, this intermediate increase was small in canola crops, meaning that ecosystem services likely did not vary strongly with distance from field boundaries (Scenario 2, Figure \@ref(fig:hypotheses)).
There was a fairly large (0.75 T/ha) intermediate increase in average wheat yields away from shelterbelts, indicating a potential reduction in ecosystem services with distance.
However, there was a similar (albeit weaker) pattern associated with bare ground, so this pattern may be caused by something other than the presence of shelterbelts.
This indicates that shelterbelts may have a positive effect on wheat yields, and future work should expand this across a larger geographic domain in order to inform growers about how best to manage the landscapes surrounding their fields.

<!-- Need a paragraph on wheat + shelterbelt effect -->

<!-- For example, average canola yield was about 0.5 T/ha lower close to Standard field edges, and yield SD was about 0.3 T/ha SD higher. -->
Why did we find few consistent effects of boundary type on crop yield, aside from shelterbelts?
Our work examined only the first and last links in the causal chain _Field boundary_ $\rightarrow$ _Insect Abundance_ $\rightarrow$ _Crop Yield_, meaning that are two mediating relationships where the overall relationship could be broken or obscured.
First, field boundaries may not have any influence on insect abundance if separated from the larger landscape context, or spillover of beneficial organisms may be low.
Second, beneficial insect abundance may not influence crop yield if pest pressure or pollination deficits are low, which may vary yearly, or by location.
<!-- However, our work does show that precision yield data could easily be joined to data on insect abundance to model this more completely. -->
Our work suggests that we need to better examine the context in which a) beneficial organisms spillover from field boundaries and b) when organisms in crops provide tangible ecosystem services.
We consider these reasons below:

### 2021-10-22 Manuscript/supplemental.Rmd
Added appendix with filter

# Appendix A: Supplementary Material {-}

```{r , echo=FALSE, fig.cap='Effect of ground speed and order (sequence of harvest) on canola yield.', out.width='100%'}  
  include_graphics('../Figures/groundSpeed_canola.png',dpi=NA)
```

```{r , echo=FALSE, fig.cap='Effect of ground speed and order (sequence of harvest) on wheat yield.', out.width='100%'}  
  include_graphics('../Figures/groundSpeed_wheat.png',dpi=NA)
```

>>>>

# Appendix A: R Code for Filtering Procedure {-}

```{r filterCode, eval=FALSE, include=TRUE}
# Helper functions -----------------

#"Inlier" spatial filtering procedure from Vega et al 2019, rewritten to work with sf + dplyr. Returns boolean
vegaFilter <- function(data,ycol,pvalCutoff=0.05,nDist=40){
  require(sf)
  require(sp)
  require(spdep)
  require(tidyverse)

  if(!any(class(data) %in% 'sf')) stop('Dataframe must be sf object')

  coords <- st_coordinates(data) #Get coordinates in matrix form

  #Get neighbourhood weights from 0 to ndist meters
  nWeights <- dnearneigh(coords,0,nDist)

  if(any(sapply(nWeights,length)==1)) warning('Some points had no neighbours and were removed')

  #Get neighbourhood indices for each point (which other points are in this point's neighbourhood?)
  nIndices <- nb2listw(nWeights, style = "W",zero.policy = TRUE)

  yield <- pull(data,{{ycol}}) #Get yield data column

  #Local Moran's I for each neighbourhood
  LM <- localmoran(yield,nIndices,p.adjust.method="bonferroni",alternative ="less")

  results <- data.frame(LM) %>% 
    rename('pval'=contains('Pr.z.')) %>% 
    mutate(keepThese= Ii > 0 | pval > pvalCutoff) %>% #Filter negative Ii and pvals < 0.05
    mutate(keepThese=ifelse(is.na(keepThese),FALSE,keepThese)) #NAs (points had no neighbours)

  ret <- pull(results)
  return(ret)
}

#Function to filter anything above a certain z-score. Returns boolean
ZscoreFilter <- function(x,z=3){
  xmean <- mean(x,na.rm=TRUE)
  xsd <- sd(x,na.rm=TRUE)
  zscore <- abs(((x-xmean)/xsd))
  zscore<z
}

#Function to filter anything above certain quantiles. Returns boolean
QuantileFilter <- function(x,quant=0.99){ 
  l <- c((1-quant)/2,1-(1-quant)/2) #Symmetric quantiles
  x>quantile(x,l[1]) & x<quantile(x,l[2]) 
}

#Function to filter large changes in bearing. Returns boolean unless returnDiffs==TRUE
bearingFilter <- function(bearing,q=NULL,z=NULL,returnDiffs=FALSE){
  if(!xor(is.null(q),is.null(z))&!returnDiffs) stop('Input quantiles or Z-score')
  #Difference in compass bearings (in degrees)
  bearingDiff <- function(x1,x2){
    x <- x1-x2
    x <- ifelse(abs(x)>180,x-(360*sign(x)),x) #Angle differences can't be >180
    return(x)
  }

  #Looks 1 point ahead and behind
  bd <- cbind(bearingDiff(lag(bearing),bearing),
              bearingDiff(lead(bearing),bearing))
  bd <- apply(bd,1,function(x) max(abs(x),na.rm=TRUE)*sign(x[which.max(abs(x))])) #Maximum bearing difference ahead and behind

  if(returnDiffs) return(bd) #Return bearing differences only, without filtering

  if(!is.null(q)){
    ret <- QuantileFilter(bd,q=q)
  } else {
    ret <- ZscoreFilter(bd,z=z)
  }
  return(ret)
}

#Positional difference filter - filters out very distant and very close points. Returns boolean unless returnDiffs==TRUE
posFilter <- function(data,q=NULL,returnDiffs=FALSE){ 
  require(sf)
  if(is.null(q)&!returnDiffs) stop('Input upper quantile')

  if(units(st_distance(data[1,],data[2,]))$numerator!='m') warning('Position differences not in meters')
  coords <- st_coordinates(data) #Get coordinates
  pdiff <- sapply(1:(nrow(coords)-1),function(i) as.numeric(dist(coords[i:(i+1),]))) #Distances between points
  pdiff <- cbind(c(pdiff,NA),c(NA,pdiff)) #Forward and backward lags
  pdiff <- apply(pdiff,1,max,na.rm=TRUE) #Maximum distance ahead and behind
  if(returnDiffs){
    return(pdiff)
  } else {
    return(QuantileFilter(pdiff,q)) #Note: uses 2-sided quantiles
  }
}

#Filter for (forward and backward) lagged speed differences. Returns boolean
dSpeedFilter <- function(speed,l=c(-1,1),perc=0.2){

  llist <- sapply(l,function(x) (lag2(speed,x)-speed)/lag2(speed,x)) #Matrix of % diffs

  #Are any lagged speed values > percent change threshold?
  ret <- !apply(llist,1,function(y) any(abs(y)[!is.na(y)]>perc))

  return(ret)
}

#Overloaded lag function that takes negative values
lag2 <- function(x,n){
  require(dplyr)
  if(n==0) {
    return(x) #No lag
  } else if(n>0){
    lag(x,n) #Positive lag
  } else {
    lead(x,abs(n)) #Negative lag
  } 
}

#Filtering pipeline  --------------------------

library(tidyverse)
library(sf)

#Note: dat == sf dataframe of yield data with columns: DryYield (T/ha), TrackAngle (compass bearing), Speed (km/hr), cropType (Wheat, Canola, Peas)

dat <- dat %>% 
  #Vega et al 2019 spatial "inlier" filter - takes a few minutes...
  mutate(vegaFilt = vegaFilter(.,DryYield,nDist = 30))

dat <- dat %>% 
  #"BS" filter - removes values >10.75 T/ha for wheat, >8 T/ha for peas & canola
  mutate(noBS = DryYield<ifelse(cropType=='Wheat',10.75,8)) %>% 
  mutate(Qfilt = QuantileFilter(DryYield,q=0.98)) %>% #Trim dry yield outliers
  mutate(bFilt = bearingFilter(TrackAngle,q=0.98)) %>% #Trim extreme bearing changes (turning)
  mutate(speedFilt = QuantileFilter(Speed,q=0.98)) %>% #Trim absolute speed outliers
  #Trim speed differences (>20% change 2 steps forward and backward, suggested by Lyle et al 2014)
  mutate(dSpeedFilt = dSpeedFilter(Speed,l=c(-2,-1,1,2),perc = 0.2)) %>% 
  mutate(posFilt = posFilter(.,q=0.98)) %>% #Trim points that are far away from eachother
```
